The Forensic Interchange Format - a file format for digital evidence

### 2009-02-21 paper/paper.tex
Updated paper to talk about hashing and encryption.

%\documentclass[10pt,twocolumn]{article}
\documentclass[12pt, conference]{IEEEtran}
%\IEEEoverridecommandlockouts
%\documentclass{article}
%\documentclass[12pt]{IEEEtran}
%\documentclass{elsart5p}

% Make sure we know its a draft for now
%\usepackage{draftwatermark}
\usepackage{verbatim}
%\documentclass[12pt]{article}
\usepackage{epsfig}
\usepackage[hypertexnames=false,bookmarksopenlevel=1,bookmarksopen,bookmarksnumbered,colorlinks,plainpages=false,linktocpage]{hyperref}
%\usepackage{natbib}
%\pagestyle{empty}
%\psdraft
%\baselineskip=20pt %Sets line spacing to 1 unit
%\bibliographystyle{elsart-num-names}
\bibliographystyle{IEEEtran}
\usepackage{cite}
\newenvironment{definition}[1][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1:}]}{\end{trivlist}}

@@

\begin{abstract}
Forensics analysis requires the acquisition of management of many
different types of evidence material. Often the same evidence is
reviewed by several different tools. We propose an open, extensible
file format for storing of evidence and interchanging analysis results
among different tools. The Forensic Interchange Format (FIF) was
designed to be simple to implement, relying on the well supported Zip
specifications for bit level file access.
\end{abstract}

@@

The Zip file specifications are publicly available and widely
implemented. The Zip64 extensions are provided to enable Zip archives
to grow beyond a two gigabyte limit by using 64 bit
references. Although Zip64 also defines encryption and authentication
extensions, we do not use these in this specification due to the
restrictions imposed on their use. Rather, we specify stream based
encryption scheme for use in FIF files.

@@

Attributes MUST be specified at the top level of each volume
(i.e. must have a segment named ``properties'')

@@

\subsection{FIF Volumes and segment merging}
There are a number of aspects in the Zip file format specifications
which we do not utilize. For example, the Zip64 specification allows
for a multitude of encryption and hashing algorithms. Some of these
features however, seem to be patent encumbered. Since the FIF
specification aims to separate high level information such as hashes,
cryptography and signatures from the low level bit level
representation we implement those features at a higher level.

In many forensic applications, individual segments within the volume
need to be updated \cite{AFF}. These updates can cause the volume to
become fragmented as holes appear within the middle of the volume, and
updated segments get added to the end.

When we wish to add a new segment to an existing volume, the new
segment header SHOULD overwrite the volume central directory, and the
central directory should be rewritten including the new entry
immediately after the last file in the archive. If it is not possible
to overwrite the file, the new segment can be written at the end,
followed by an updated copy of the central directory (in that case the
old central directory remains in the file). This is illustrated in
Figure \ref{zip_structure}.

@@

\begin{figure}[tb]
  \begin{center}
  \mbox{\epsfxsize=0.6\columnwidth \epsffile{usage1.eps}}
  \caption{A typical usage scenario. Both Alice and Bob receive a FIF
volume 1 but work independently. Rather than modifying Volume 1, they
each create another volume and save their results to the new
volume. They can now exchange the new volume and effectively merge
their results into the same FIF set.}
  \label{usage}
  \end{center}
\end{figure}

This is made possible because each volume is independent of one
another, and can be loosely collated into a bigger FIF file
set. However, considering the above scenario, is it clear that
although volume 1 can be used by itself, both volume 2 and 3 might
refer to data in volume 1 and require it loaded too. We specify a
volume attribute named ``volume'' which links to the location of other
volumes.

@@

A stream is a high level abstraction of a single source of data. For
example, a stream may contain a disk image or a network
capture. Although we do not specify the exact API, implementations
MUST make a stream opened for reading have at a minimum a read, seek
and tell methods. Seeking MUST support the whence argument as
described in the standard C implementation.

Streams opened for writing MAY only support the write method as
seeking in compressed streams may not be possible. Applications should
avoid needing to seek in output streams, and try to structure writes
serially. It is usually possible to reorder a sequence of non
overlapping random writes into a linear stream and a map which is
described below.

All streams within the FIF file MUST have a properties segment. The
properties segment describes the stream attributes as specified
previously. Streams must have the following mandatory attributes:

\begin{itemize}
\item {\em type} is the kind of stream.

\item {\em size} is the size of the stream in bytes.
\end{itemize}

The type attribute is used to interpret the other segments within the
stream in a stream specific way.

@@

\subsection{The Image Stream}
The {\em Image} stream contains a single read only forensic data
source. For example, this Stream might contains a hard disk image, a
memory image or a network capture (in PCAP format). The Image stream
provides random read access into the stream (i.e. allows for randomly
seeking the stream to any point).

@@

%% Talk about signing and hashing here....

\subsection{The Map Stream}
Linear transformations of data are commonplace in forensic
analysis. For example, a file is often simply a collection of bytes
drawn from an image. A TCP/IP stream is simply a collection of
payloads from selected network packets. Sometimes the same data may be
viewed in a number of ways - for example a Virtual Address Space is a
mapping of the Physical Address Space through a page table
transformation. It is, therefore, beneficial to define a way in which
the underlying storage system can apply transformations to existing
data to produce transformed data.

@@

\subsection{The Overlay stream}
The Overlay stream is a variant of the {\em Image} stream in that it
allows for images to be stored in (possibly compressed) chunks. The
main utility of the Overlay stream is in providing transparent access
to other forensic formats, such as EWF, AFF. These formats also use
chunk based compression to provide the ability to randomly seek while
reading. To access these file formats we only need to have an index of
each chunk offset, and then directly use the overlayed files. The {\em
ewf2fif} tool provides such an overlay feature for EWF volumes by
employing libewf to scan the internal data structures.

@@

\begin{figure}[tb]
  \begin{center}
  \mbox{\epsfxsize=0.8\columnwidth \epsffile{overlay.eps}}
  \caption{The use of an Overlay Stream to overlay an image with
coalesced chunks.}
  \label{overlay}
  \end{center}
\end{figure}

@@

\subsection{The Private stream}

\subsection{Encrypted Streams}

@@

>>>>

%\documentclass[10pt,twocolumn]{article}
\documentclass[10pt, conference]{IEEEtran}
%\IEEEoverridecommandlockouts
%\documentclass{article}
%\documentclass[12pt]{IEEEtran}
%\documentclass{elsart5p}

% Make sure we know its a draft for now
\usepackage{draftwatermark}
\usepackage{verbatim}
%\documentclass[12pt]{article}
\usepackage{epsfig}
\usepackage[hypertexnames=false,bookmarksopenlevel=1,bookmarksopen,bookmarksnumbered,colorlinks,plainpages=false,linktocpage]{hyperref}
%\usepackage{natbib}
%\pagestyle{empty}
%\psdraft
%\baselineskip=20pt %Sets line spacing to 1 unit
%\bibliographystyle{elsart-num-names}
\bibliographystyle{IEEEtran}
\usepackage{cite}
\newenvironment{definition}[1][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1:}]}{\end{trivlist}}

@@

\begin{abstract}
Forensics analysis requires the acquisition and management of many
different types of evidence material. Often the same evidence is
reviewed by several different tools. We propose an open, extensible
file format for storing of evidence and interchanging analysis results
among different tools. The Forensic Interchange Format (FIF) was
designed to be simple to implement, relying on the well supported Zip
specifications for bit level file access.
\end{abstract}

@@

The Zip file specifications are publicly available and widely
implemented. The Zip64 extensions are provided to enable Zip archives
to grow beyond a two gigabyte limit by using 64 bit
references. Although Zip64 also defines encryption and authentication
extensions, we do not use these in this specification due to the
restrictions imposed on their use. Rather, we specify stream based
encryption scheme for use in FIF files (See section
\ref{crypted_stream}).

@@

Attributes MUST be specified at the top level of each volume
(i.e. must have a segment named ``properties''). The volume attributes
are used to specify the volume UUID.

@@

\label{target_reference}
{\em A target reference} is a way of referencing objects. This is used
in a number of places which will be highlighted below. Using
references it is possible for the FIF file to refer to objects
embedded within itself, or within other FIF files. Its is also
possible to refer to external flat files. A number of usage scenarios
are described in Section \ref{usage_scenarios}. The reference is
specified with the following conventions:

\begin{itemize}
\item A reference starting with {\em file://} refers to an external
file. The external file reference must be sanitised (to remove
.. directory traversal) and rooted at the current directory.

\item A reference starting with {\em http://} refers to a HTTP
object which will be fetched. Support for this is optional.

\item A reference in the format {\em fif://UUID/stream/segment} refers
to an external FIF file set (identified by its UUID). If the segment
is missing (i.e. the reference ends with a ``/''), the reference is
for the entire stream. We do not specify a way of actually resolving
the UUID into a physical locations. This can achieved by
implementations using many methods, such as Active Directory, LDAP,
DNS, flat files or SQL databases.

\item A references ending with a ``/'' signify a reference to a
stream in the current FIF file. The stream will be opened in its
entirety. If the reference does not end with a ``/'', the reference is
for a single segment.
\end{itemize}

\subsection{FIF Volumes and segment merging}
There are a number of aspects in the Zip file format specifications
which we do not utilize. For example, the Zip64 specification allows
for a multitude of encryption and hashing algorithms. Some of these
features however, seem to be patent encumbered. Since the FIF
specification aims to separate high level information such as hashes,
cryptography and signatures from the low bit level representation, we
implement those features at a higher level.

In many forensic applications, individual segments within the volume
need to be updated \cite{AFF}. These updates can cause the volume to
become fragmented as holes appear within the middle of the volume, and
updated segments are added at the end.

When we wish to add a new segment to an existing volume, the new
segment header SHOULD overwrite the volume central directory, and the
central directory should be rewritten including the new entry
immediately after the last file in the archive. If it is not possible
to overwrite the file, the new segment can be written at the end,
followed by an updated copy of the central directory (in that case the
old central directory remains in the file). This is illustrated in
Figure \ref{zip_structure}. Implementations may choose to delay
writing the central directory until the file is closed.

@@

\begin{figure}[tb]
  \begin{center}
  \mbox{\epsfxsize=0.6\columnwidth \epsffile{usage1.eps}}

  \caption{A typical usage scenario. Both Alice and Bob receive a FIF
  volume but work independently. Rather than modifying the volume,
  they each create another volume and save their results to the new
  volume. They can now exchange the smaller new volumes and
  effectively merge their results into the same FIF set.}

  \label{usage}
  \end{center}
\end{figure}

This is made possible because each volume is independent of one
another, and can be loosely collated into a bigger FIF file
set.

@@

A stream is a high level abstraction of a single source of data. For
example, a stream may contain a disk image or a network
capture. Although we do not specify the exact API, implementations
MUST make a stream opened for reading have at a minimum a read, seek
and tell methods. Seeking MUST support the whence argument as
described in the standard C implementation (i.e. allow seeking to the
end of the stream).

Streams opened for writing MAY limit support for seeking, as efficient
seeking in compressed streams may not be possible. Applications should
avoid needing to seek in output streams, and try to structure writes
serially. It is usually possible to reorder a sequence of non
overlapping random writes into an Image stream and a Map Stream (These
are discussed in Sections \ref{image_stream} and
\ref{map_stream}). Implementations are free to hide the inability to
seek in writing streams by automatically implementing such a scheme,
but this is not mandated by the specification.

All streams within the FIF file MUST have a properties segment. The
properties segment describes the stream attributes as specified
previously. Streams must have the {\em type} attribute which allows
implementations to determine how they should be processed.

@@

\subsection{The Image Stream}
\label{image_stream}
The {\em Image} stream ({\em type=image}) contains a single read only
forensic data source. For example, this Stream might contains a hard
disk image, a memory image or a network capture (in PCAP format). The
Image stream provides random read access into the stream (i.e. allows
for randomly seeking the stream to any point). The {\em size}
attribute MUST be specified to determine how large the stream is.

@@

Chunks can optionally be individually hashed, allowing for detection
of modifications. The attribute {\em hash\_type} specifies the type of
hash used (implementations SHOULD support md5,sha1 and sha256, but can
define others). The hashes are stored in any number of segments within
the stream, of arbitrary name. The hash segment names is provided by
using the {\em hash} stream attribute.

Each hash segment MUST be a list of a chunk id encoded in 32 bit
little endian integer format followed by the hash (the length of which
depends on the hash itself). This approach is illustrated in Figure
\ref{hash_index}.

Hash segments can contain any number of chunk hashes in them. It is
recommended that a hash segment be created when a volume is closed to
cover all the chunks within that volume. This allows for hashes to be
verified even if some volumes are missing. It is permitted for a hash
segment to update a hash of a chunk previously set by an earlier hash
segment - this may be required when a chunk is updated. The hash
segment can then be signed to assure the integrity of the updated
segment.

\begin{figure}[tb]
  \begin{center}
  \mbox{\epsfxsize=0.6\columnwidth \epsffile{hash.eps}}

  \caption{The structure of a hashing segment. The attributes {\em
  hash} and {\em hash\_type} declare the hash segment. The segment
  contains a list of integer encoded chunk id followed by an unencoded
  hash for each chunk.}

  \label{hash_index}
  \end{center}
\end{figure}

This design differs from AFF, which creates a separate segment for
each hash. The Zip headers are more verbose than the AFF segment
headers, and typically FIF chunks are much smaller (default 32kb, vs
16Mb for AFF). Therefore, creating a new segment for each hash value
results in more overheads in the FIF format. By packing the hashes
into a few segments it is possible to reduce this overhead to a
minimum.

Hashes are calculated in an identical way to that defined by AFF,
thats is the full name of the segment (including stream component) is
followed by a single NULL byte, followed by the (uncompressed) content
of the chunk. The inclusion of the segment name ensure that even
identical chunks in content will have a different hash.

\subsection{The Map Stream}
\label{map_stream}
Linear transformations of data are commonplace in forensic
analysis. For example, a file is often simply a collection of bytes
drawn from an image. A TCP/IP stream is simply a collection of
payloads from selected network packets. Sometimes the same data may be
viewed in a number of ways - for example a Virtual Address Space is a
mapping of the Physical Address Space through a page table
transformation. It is, therefore, beneficial to define a way in which
the underlying storage system can apply transformations to existing
data to produce transformed data.

@@

\subsection{The Overlay stream}
The Overlay stream is a variant of the {\em Image} stream in that it
allows for images to be stored in (possibly compressed) chunks. The
main utility of the Overlay stream is in providing transparent access
to other forensic formats, such as EWF or AFF. These formats also use
chunk based compression to provide the ability to randomly seek while
reading. To access these file formats we only need to have an index of
each chunk offset, and then directly use the overlayed files. The {\em
ewf2fif} tool provides such an overlay feature for EWF volumes by
employing libewf to scan the internal data structures.

@@

\begin{figure}[tb]
  \begin{center}
  \mbox{\epsfxsize=0.6\columnwidth \epsffile{overlay.eps}}
  \caption{The use of an Overlay Stream to overlay an image with
coalesced chunks.}
  \label{overlay}
  \end{center}
\end{figure}

@@

\subsection{The Private stream}
Often forensic software needs to store internal state, such as the
current state of the GUI, or internal data structures which may be
required for caching. Usually, software create additional cache files
to maintain this information. It is advantageous to be able to store
these in the evidence file itself. The evidence file can then be
re-opened by the software at a later stage and private data can be
retrieved and used directly. Such private application data can be
stored in a Private stream. There is no specification of what can be
stored in the private stream, and applications with private streams
should be able to store arbitrary segments in any format at all.

It is recommended that applications name their private segment
sufficiently accurately so as not to be confused with other
applications, or different versions of the same application
(e.g. ``pyflag/0.87/cache/'').

\subsection{Encrypted Streams}
Encryption is an important property in an evidence file format
\cite{AFF}. In particular with the FIF format, multiple streams may be
present in the file set, and often different access levels are
desired. For example, for evidence set containing both network
captures and disk images it may be desirable to limit access to
streams based on legal authorizations.

Although the Zip64 standard specifies encryption, it is not suitable
for our purposes since it encrypts each segment separately, and does
not specify a sufficiently flexible scheme (e.g. support for PKI, PGP
keys). Segment based encryption may lead to information leakage when
segments are compressed, as the uncompressed size of the segment may
be deduced.

The Encrypted Stream ({\em type=encrypted}) SHOULD be implemented by
FIF implementations, but may be omitted with an obvious loss of
functionality (Encryption may be prohibited or restricted in some
jurisdictions).

The Encrypted Stream is similar to the Image Stream above in that data
is written out in chunks (although its probably not effective to
compress these chunks at all). Each chunk is to be encrypted upon
writing and transparently decrypted using a scheme declared in the
{\em scheme} attribute.

The Encrypted Stream may contain any data at all. It is useful
however, to store an actual FIF volume within the Encrypted
stream. This provides block level encryption for the contained FIF
volume. It is recommended that Encrypted Streams containing a FIF
volume set an attribute ({\em content-type=application/x-fif-volume})
to allow implementations to automatically recognize volumes. This
approach is illustrated in Figure \ref{crypted_fif}.

\begin{figure}[tb]
  \begin{center}
  \mbox{\epsfxsize=0.8\columnwidth \epsffile{crypted.eps}}
  \caption{Embedding an encrypted FIF volume within an Encrypted Stream.}
  \label{crypted_fif}
  \end{center}
\end{figure}

The result is that a number of FIF volume are used as {\em Container
Volumes} to provide storage for Encrypted Streams. The main {\em
Embedded Volume}, which actually contains data is stored within the
Encrypted Stream, effectively distributed throughout the container
volumes. Note that the Container Volume may contain several Encrypted
Streams and therefore contain multiple Encrypted FIF
Volumes. Container Volumes may contain non encrypted streams as well,
and may implement different encryption schemes and keys for each
Encrypted scheme. This effectively allows arbitrary access policies to
be implemented as only volumes which can be accessed can be merged.

Implementations should create embedded FIF volumes as extensions to
the container volumes (i.e. with the same UUID). Thus when embedded
volumes are opened, their streams become automatically viewable as
part of the evidence set.

\subsection{Encryption Schemes}
We specify a number of encryption schemes, but extra schemes may be
defined by the implementation.

\subsubsection{Null Encryption}
This scheme ({\em scheme=null}) specifies no encryption at all. This
may be useful for testing an implementation, but obviously provides
no real security.

\subsubsection{AES-SHA-PSK}
This scheme ({\em scheme=aes-sha-psk}) uses AES for block level
encryption, with the following SHA based ESSIV scheme, and Pre-Shared
Key (passphrase).

When creating a new volume, a master key is generated using the first
128 bits of a SHA1 hash of the Pre-Shared-Key appended to a 64 bit
random salt:
\begin{eqnarray}
Key_{Master} = \left | SHA1(salt + PSK) \right | _{128}
\end{eqnarray}

For each chunk, the chunk IV is obtained by taking the first 128 bits
of the SHA1 hash of the chunk id encoded as a 32 bit little endian
integer, followed by the master key:
\begin{eqnarray}
IV = \left | SHA1(chunk\_id + Key_{Master}) \right | _{128}
\end{eqnarray}

The block is then encrypted or decrypted using AES with this IV.

Implementations MUST specify the salt as a stream attribute named {\em
salt} in base 64 encoding. Implementations are free to specify any API
for passing the Pre-Shared Key, for example the PSK can be specified
in a configuration file, typed in at the terminal, passed in using
process environment variables, or encoded within the FIF filename
itself.

\subsection{Optional features}
Implementations may implement the following features to simplify the
interfaces.

\subsubsection{Automated volume loading}
FIF volumes are effectively independent, as each volume does not need
to know anything about other volumes. This makes it necessary to
specify in advance all volumes for the implementation to merge them
all. In particular it may be necessary to specify encrypted volumes
(which may not be read if the user has no authorization for the
Encrypted Stream).

As an optional enhancement FIF specifies a volume attribute named {\em
volume} (i.e. an attribute present in the top level ``properties''
segment). This is a target reference (see section
\ref{target_reference}) for other volumes. 
Implementations should attempt to open external references (e.g. using
the {\em file://} convention) before internal references. As new
volumes are merged their volume attributes must be scanned for
additional references. Each volume must list all other known volumes
as volume attributes.

In this way, encrypted volumes, and external volumes can be located
and found automatically.

@@

\section{Usage Scenarios}
In this section we describe how a FIF file format may be used in
various situations. Many of those can be solved using other forensic
file formats, but often in a more awkward way.

\subsubsection{Rapidly converting a set of DD images}
Many hardware devices are available to acquire hard disks in the
fields. These often produce a set of uncompressed images split at a
certain size. If the images are split at exactly the same size its
possible to create a FIF archive consisting of a single Overlay
Stream, with a chunk\_size set to the size of each of the images. The
images can then be listed in order as external targets (e.g. {\em
target=file://disk1.dd}). This overlay can be created instantly
without needing to re-compress any of the images. In addition, digital
signatures can be generated for the whole stream or for each segment.

\subsubsection{Rapid conversion of EWF file sets}
Much existing evidence has been acquired using the EWF file
format. This format, similarly, consists of a series of chunks which
may be compressed, with interdispersed indexes. A tool such as
``ewf2fif'' is able to utilize the libewf library to create an Overlay
stream of the original image without the need to re-compress it. This
conversion is very quick and the resulting FIF file is very small -
each volume in the EWF file set is referenced through an external
target reference, and the Overlay Stream is built as an index of each
chunk location.

\subsubsection{Acquisition of RAID disks}
Often disks in a system are grouped into RAID devices, commonly RAID-5
or RAID-0. Previously, if disks were acquired independently, they
would need to be analysed using a tool which was able to reassemble
RAID devices.

With the FIF format, each of the disks can be acquired as a separate
Image Stream. Finally a tool such as PyFlag may be used to deduce the
RAID map, which can be appended to the FIF file as a Map Stream. This
Map Stream can then be opened by any tool without the tool needing to
have explicit support for RAID reassembling.

\subsubsection{Cryptographic management of evidence}
A FIF archive may hold multiple encrypted volumes, each in its own
Encrypted Stream. Each of those streams is encrypted using a different
master key, and therefore can have different passphrases, and can be
assigned to different users by encrypting the master key with
different X509 certificates. Its is also possible for users to create
non-encrypted volumes within the FIF file.

This can be used to enforce access controls in line with current
legislative requirements. For example, within the same investigation
different material is often obtained under different warrants
(e.g. wiretap authorizations are different from search
warrants). Therefore, different investigators and analysts need
different access to the different streams. However, the analysts may
still store the results of their analysis in an unencrypted form, or
assign others permissions to decrypt their analysis results, without
providing access to the underlying data.

This can be used in sharing meta data (e.g. Map Streams of files of
interest) between analysts, without needing to provide access to the
underlying data.

\subsubsection{Forensic Application State}
Often forensic applications need to store files other than the
evidence itself - for example, they might need to store internal data,
annotations, keyword hits etc. Currently these applications store the
data in a proprietary external file or database. This makes it
difficult to archive because the evidence itself may become separated
from the case file.

It is advantageous for these applications to store their state within
the evidence file itself using a Private Stream. Then when opening the
evidence file again, the results of their analysis will become
available.
